---
title: "Evolutionary Scale Mode (ESM)"
author: 
  - Nicholas P. Cooley, Department of Biomedical Informatics, University of Pittsburgh
date: "`r Sys.Date()`"
output:
  github_document: default
always_allow_html: true
---

# This is a work in progress...

Performing large scale structure prediction is a useful thing but resource usage and management are non-trivial for tasks that require big GPUs, particularly when using distributed and scavenged resources like those provided by the Open Science Pool.

## First things first

Deploying ESM on the OSPool relies, to some extent, on a static version of a model that can be distributed in a manner that gives a user assurance that the model is not changing behind the scenes as they're pulling it from wherever it lives. Currently ESM has an open model maintained on huggingface.

So some kind of staging of the model you want to use will generally be an initial step. If you want to do this in a docker container, you just need to ensure you mount a directory to send the model back to your local environment.

```{bash, echo = TRUE, eval = FALSE}
# pull a model to the docker container and dump it back to a local directory
# step one: spin up your docker container
docker run -it --rm -v <some_local_dir>:<some_container_directory> npcooley/esm:0.0.4
```

Get the model - either interactively in a python REPL or through a script. This requires a token.

```{python, python.reticulate = FALSE, echo = TRUE, eval = FALSE}
# pull a model to the docker container and dump it back to a local directory
# step two: get the model, this is just directly from the ESM folks' tutorials

from huggingface_hub import login
from esm.models.esm3 import ESM3
from esm.sdk.api import ESM3InferenceClient, ESMProtein, GenerationConfig

# Will instruct you how to get an API key from huggingface hub, make one with "Read" permission.
login()

# This will download the model weights and instantiate the model on your machine / in the container if you're using that route.
# My local machine does not have a GPU :(
model: ESM3InferenceClient = ESM3.from_pretrained("esm3_sm_open_v1").to("cpu") # or "cuda"
```

Create a tarball of the cache, move it to the mounted volume and you will now be able to distribute the tarball as you would any normal large file within the OSPool.

```{bash, echo = TRUE, eval = FALSE}
# pull a model to the docker container and dump it back to a local directory
# step three: as of the creation of this document, the model gets stored in '$HOME/.cache'

tar czvf cache.tar.gz ${HOME}/.cache/huggingface # bog standard
# or if we're really worried about space -- disk is cheap on the OSG, but not costless, time and memory are always your biggest pain points
tar cJvf cache.tar.xz ${HOME}/.cache/huggingface # takes a lot of time, but a smaller tarball
```

From here, you should just be able to unpack this tarball into your `.cache` directory, and ESM will find it without too much hassle. It is good practice to scrub your token from the huggingface directory in `.cache` before you build this tarball. And of course, you could do all this in your local environment without using containers.

# Example 1 local testing:

## Scripts

A relatively simply python script gives us access to structure prediction, and can be made slightly more complicated to return some observations about resource usage.

```{bash}
cat local_testing/GetStructs.py
```

Another relatively simple script that can be called inside our docker container locally to produce some data.

```{bash}
cat local_testing/UsageExample.sh
```

## Example Data

I typically prefer to keep my local environment out of any work that eventually needs to be replicable or exist in a form that can be referenced easily for future variations. The following R code, when executed by the knitr engine will just place some test sequences in the working directory of this `.Rmd` file. Which is then easily referenced by both this document and docker calls below. The `local_testing` folder is present in this repo, though without the `cache.tar.gz` file (it's not a small file...).

```{r}
suppressMessages(library(DECIPHER))
# a fairly randomly selected example stringset
x <- "ftp://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/023/585/805/GCF_023585805.1_ASM2358580v1/GCF_023585805.1_ASM2358580v1_protein.faa.gz"
x <- readAAStringSet(x)

# smallest seqs
y1 <- head(x[order(width(x))], n = 15)
writeXStringSet(x = y1,
                filepath = "local_testing/testseqs1.fa")

# randomly selected seqs
set.seed(1986)
y2 <- sample(x, size = 15)
writeXStringSet(x = y2,
                filepath = "local_testing/testseqs2.fa")

# biggest seqs
y3 <- tail(x[order(width(x))], n = 15)
writeXStringSet(x = y3,
                filepath = "local_testing/testseqs3.fa")
```

## Docker Stuff

Spin up our docker container:

```{bash}
# -dit
# d == detached state
# i == interactive, leave stdin open even AFTER the main process stops
# t == simulate terminal interface
docker run --name ESM3 \
  -v ./local_testing:/tempdir \
  -dit npcooley/esm:0.0.4
```

Run our scripts. This is a little bit of turtles all the way down, but functionally, we're just using bash scripts specifically written for the use case of this example to call our python script (and some housekeepers) that is written to be more generally applicable. All inside our docker container which is a static environment while using a *somewhat ambiguously named* cached version of a publicly available model.

```{bash}
time docker exec ESM3 tempdir/UnpackModel.sh > log.txt

time docker exec ESM3 tempdir/UsageExample.sh testseqs1.fa >> log.txt 2>&1
time docker exec ESM3 tempdir/UsageExample.sh testseqs2.fa >> log.txt 2>&1
time docker exec ESM3 tempdir/UsageExample.sh testseqs3.fa >> log.txt 2>&1

docker exec ESM3 tempdir/PackageResults.sh
```

We can then examine our resource usage from the data we've generated. Because of limitations with memory tracking *in python* some memory values may be negative, and in general these values may not be the most accurate. R will throw a warning for that while trying to log the x-axis if it happens, and in this case it's fine.

```{r}
suppressMessages(library(yaml))
#| dev = c('png', 'pdf'), fig.width = 7, fig.height = 3.5, fig.align = "center",

untar(tarfile = "local_testing/etc.tar.gz", exdir = tempdir())

files01 <- list.files(path = tempdir(),
                      pattern = "WP_.+\\.txt",
                      full.names = TRUE)

tab1 <- vector(mode = "list",
               length = length(files01))
for (m1 in seq_along(files01)) {
  tab1[[m1]] <- read_yaml(files01[m1])
}

tab1 <- do.call(rbind,
                tab1)
tab1 <- data.frame("ID" = unlist(regmatches(m = gregexpr(text = files01,
                                                          pattern = "WP_[^.]+"),
                                             x = files01)),
                   "mean_pLDDT" = as.numeric(tab1[, 1]),
                   "pTM" = as.numeric(tab1[, 2]),
                   "seq_len" = as.integer(tab1[, 3]),
                   "time" = as.numeric(tab1[, 4]),
                   "mem_in_MiB" = as.numeric(unlist(regmatches(m = gregexpr(pattern = "^[^ ]+",
                                                                            text = tab1[, 6]),
                                                               x = tab1[, 6]))))

layout(mat = matrix(c(1,2),
                    nrow = 1))
par(mar = c(4,3,3,0),
    mgp = c(2, .75, 0))
plot(y = tab1$seq_len,
     x = tab1$mem_in_MiB,
     pch = 20,
     log = "x",
     ylab = "Sequence length (AA)",
     xlab = "Memory footprint (MiB)",
     col = ifelse(test = tab1$mean_pLDDT > 0.8 &
                    tab1$pTM > 0.8,
                  yes = "blue",
                  no = "red"))
par(mar = c(4,1.5,3,1.5),
    mgp = c(2, .75, 0))
plot(y = tab1$seq_len,
     x = tab1$time,
     pch = 20,
     log = "x",
     xlab = "Time (sec)",
     col = ifelse(test = tab1$mean_pLDDT > 0.8 &
                    tab1$pTM > 0.8,
                  yes = "blue",
                  no = "red"),
     yaxt = "n")
legend("topright",
       legend = c("acc.",
                  "inacc."),
       col = c("blue",
               "red"),
       pch = 20)
mtext(text = "resources by peptide size",
      outer = TRUE,
      line = -2)
```

Stop and remove our container to keep our workspace nice and tidy.

```{bash}
docker stop ESM3
docker rm ESM3
```

## Final Thoughts

As is stated above this readme is a bit of a work in progress. It is important to note that this script is testing model performance *on a cpu*, and time performance will be considerably better on almost any gpu. If I can find the time, I'll include the *results* of some tests on the OSG, though that code cannot be automatically triggered through the knitr engine like the code included in this document.

```{r}
sessionInfo()
```
